
# Introduction 

The `rviewgraph` package provides functions for creating, changing,
and viewing a graph. It originally contained only one function called
`rViewGraph()` that allowed viewing a fixed graph using an animated 
graphical user interface (GUI). 
A newer function called 'vg()` is now available and
allows for viewing a graph the structure of which can change dynamically. 
To illustrate this we will implement a simple Markov chain Monte
Carlo (MCMC) sampler with a given target distribution.

# Creating the graph viewer

First ensure that `rviewgraph` is available, and attempt to start
both an old version of the GUI using `rViewGraph()`, and new version of the GUI
using `vg()`.
For the old version we need to specify a graph in advance so 
we generate some random edge end points.

```{r setup}
library(rviewgraph)

oldgui = rViewGraph(sample(1:10,10,TRUE),sample(1:10,10,TRUE))
is.null(oldgui)
if (interactive() && !is.null(oldgui))
{
	oldgui.stop()
	oldgui.hide()
}

v = vg()
is.null(v)
```

We see that `rViewGraph()` returns `NULL` and prints an error message. 
This is because the old programs can only be run in interactive mode.
The new version, `vg()`, however, will return a list of functions that
can be run non-interactively, but, of course, without a visible GUI.
As this vignette was produced non-interactively, there is no GUI,
but if you are replicating it from a live `R` session 
you should see a new `Java` window that has appeared with
a blank canvas on which the graph will be viewed.
There will also be a GUI showing the random graph generated in the
call to `rViewGraph()`, so we remove that. 

The returned value `v` is a list of functions that we will use to
display the graph as we run the sampler.

# Defining the elements of the sampler

First we define a probability mass function
on graphs, in fact we need only specify this up to a constant
of proportionality. 
This simple example depends only on the number of edges in the graph, and
as such it would be more efficient to track this count as we iterate
through the MCMC scheme, however, this will illustrate how the structure
of the graph can be queried, and also makes the MCMC scheme more easily
adapted to different probability distributions.
For numerical stability, we work with logs of probabilities.

This probability distribution favours more sparsely edged graphs if
we set `a > 0`, and more densely edged ones if `a < 0`.
We can query the graph for a two dimensional array that specifies the
edges. Because this graph is undirected, each edge is returned
twice, once in each direction, so we divide to get the number of edges.

```{r prob}
edges = function(g)
{
	dim(g$getEdges())[1] / 2
}

logProb = function(g,a)
{
	-edges(g)*a
}
```

Here is a function that flips an edge in a graph. That is, if a pair of
vertices are connected it disconnects them, if disconnected, it connects
them. This will be our Metropolis proposal scheme, and it is reversible.
The argument `g` is assumed to be a list that was returned by 
calling `vg()`.

```{r flip}
flip = function(g,x,y)
{
	if (g$connects(x,y))
		g$disconnect(x,y)
	else
		g$connect(x,y)
}
```

Now we define a Metropolis MCMC proposal and acceptance step.
The calls to `Sys.sleep()` just slow the process down to make the
process easier to follow in the viewer when used interactively. 
The colour changes will show the proposals being made.

```{r step}
step = function(g, a, t = 1)
{
	# Find the probability of the current graph.
	delta = logProb(g,a)

	# Choose a pair of random vertices to flip, and display them.
	x = sample(g$getVertices(),2,replace=FALSE)
	g$colour(x,"red")
	Sys.sleep(t/3)
	
	# Then make the flip.
	flip(g,x[1],x[2])
	Sys.sleep(t/3)

	# Find the ratio of the probabilities of the proposed new graph
	# and the previous incumbent.
	delta = logProb(g,a) - delta

	# Then with probability equal to the minimum of 1 and this ratio 
	# accept the new graph by doing nothing, otherwise reject it by 
	# undoing the flip.
	if (log(runif(1)) > delta)
		flip(g,x[1],x[2])

	# Reset the colours, and pause before the next step.
	g$colour(x)
	Sys.sleep(t/3)
}
```

# The complete Metropolis MCMC scheme

We can now combine these functions to make a complete MCMC scheme
using here a graph of `n = 20` vertices. 
As an example of an output, we can calculate the mean number of
edges over the simulation.
If we set the penalty to 0, we get a random walk over graphs
of size `n`, so the simulated mean number of edges should be about `n(n-1)/4`.

If this is run in interactive mode, this is primarily to illustrate
how the GUI shows the progress of the sampler, so we set the
sleep parameter to give one step per second and make 50 updates.
If run non-interactively, we don't need to slow down the sampler
to see the progress, so we make more iterations and include a burn
in period to allow the sampler to move away from the initial state
and obtain better estimates of the mean number of edges.

The scheme is run using the viewer that was created at the beginning of 
this vignette.

```{r}
	# To make the simulation reproducible.
	set.seed(1)

	n = 20
	v$add(1:n)

	pen = 0

	if (interactive())
	{
		wait = 1
		s = 50
	} else
	{
		wait = 0
		s = 1000
		for (i in 1:s)
			step(v,pen,wait)
	}

	medge = 0
	for (i in 1:s)
	{
		step(v,pen,wait)
		medge = medge + edges(v)
	}
	c(medge/s, n*(n-1)/4)
```

A positive penalty should decrease the mean number of edges.

```{r}
	pen = 1
	medge = 0
	for (i in 1:s)
	{
		step(v,pen,wait)
		medge = medge + edges(v)
	}
	medge/s
```

And a negative one should increase the mean number of edges.

```{r}
	pen = -1
	medge = 0
	for (i in 1:s)
	{
		step(v,pen,wait)
		medge = medge + edges(v)
	}
	medge/s
```

